Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-11 Manuscript/manuscript.tex
Roughly skeched analysis with residence times of cations in stoichimetry section.

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
often used as model systems to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. A direct comparison of
ion binding affinities to PC bilayers was presented between simulations and
experiments by using the electrometer concept, which is based on the experimental NMR data
for the lipid headgroup order parameters~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including the electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
As a starting point we use the parameters from the Lipid14 model \cite{dickson14},
which gave the best cation binding behaviour in the previous study \cite{catte16}.
The developed ECC-lipid parameters reproduce the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly for reproducing the membrane binding affinities and induced
structural effects of Sodium and Calcium ions.

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of the lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased back
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in the Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules. \\
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models. \\
SAMULI: There is now a paragraph in the conclusions, which mentions this topic.}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stronger electrostatic interactions with them.
After reducing the $\sigma$ parameters by a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
back in agreement with the experimental value (see Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations by 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid~\cite{seelig87}.
The change of order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

@@

The relation between bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with bound positive charge and increase with bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt toward water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured order parameter change depends on both, the head group response 
to the bound charge and the amount of bound charge, i.e. $m_i$ and $X^\pm$
in Eq.~\ref{OPchangeEQ}, respectively. The former property has to be
well quantified before the electrometer concept can used to 
analyze the binding affinities, as done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response to the bound charge in simulations,
we use the experimental data for dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
The early experimental electrometer concept data for the lipid head group
order parameter changes was reported as a function 
of the salt concentration in water before solvating
the lipids~\cite{akutsu81}. The later study used atomic absorption
spectroscopy and reported the order parameter changes as
a function of the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Despite of the measurable differences between these two concentrations for CaCl$_2$ systems,
the qualitative conclusions in this or in the previous work~\cite{catte16} are
not affected.

To quantify the ion binding affinity to a lipid bilayer, we
calculated the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
This quantity was chosen because it does not depend on
%the definition of the interfacial region or
the position of the Gibbs dividing plane between two bulk regions.
%\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
%SAMULI: Here is a new trial.
Here we assume that the interface locates between the hydrophobic interior
of a lipid bilayer and the bulk water region far from the membrane.
The bulk concentration of ions and water is zero inside the bilayer.
The concentrations in bulk water region can be calculated from 
the farthest point from both lipid leaflets in the water phase.
The region between these boundaries contains all the ions and water
molecules in the simulation box.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

The structure of lipid bilayers in simulations without ions were validated against NMR and 
\mbox{x-ray} scattering experiments by calculating order parameters
for C-H bonds and form factors. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules by using Eq. \ref{OP}. Form factors were calculated 
from equation \todo{As Hector suggested, it might be better to write the simpler
form for this equation.}
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length in the bulk region,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning until the water bulk region, $D$.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
Simulations of POPC bilayer in pure water or in varying salt concentrations
contained 128 POPC molecules and approximately 50~water molecules per each lipid
in the periodic orthorhombic simulation box.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model
\todo{More justification for the choosing the OPC3 water model are needed.
It might be good to show the comparison with the scattering data in bulk water in SI.}. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
\todo{The normal TIP3P was tested as well, right?} presented in Supporting Information (SI). 
We used the ECC-ions model for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium, which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
%Chloride counter ions were used
%in simulations, because parameters for bromide, which was used in the experiment \cite{scherer89},
%were not available in the standard Gromacs files for Amber force field.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.

@@

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
The dominant contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Such a finding is supported by the calculated
number of contacts between \ce{Ca^{2+}} and POPC oxygen atoms.
The average number of contacts between \ce{Ca^{2+}} and any oxygen atom in POPC is 59.8, 
whereas if only phosphate oxygen atoms are considered 
the average number of contacts decreases only by a tiny amount to 57.2. 
We corroborate this finding with the probability isodensity contours in Fig.~\ref{fig:volmaps}, 
which are easier to interpret.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From recent theoretical work with long enough simulations this time can be roughly estimated
in the order of $1$--$10\,\mu\mathrm{s}$~\cite{javanainen17}. 
This is in contrast to our model, ECC-lipids, 
which gives a mean residence time in the order of $10\,\mathrm{ns}$, 
\todo{evaluate this number, mean residence time, accurately based on the contacts data.}
i.e. at least two orders of magnitude lower than previous estimates.
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent.

\todo{Finalize stoichiometry analysis for \ce{Na^+}, \ce{Ca^{2+}},
  their interaction energies with the lipid membrane, etc,
  and finalize the discussion after these results.}

%It is also suggested that the addition of \ce{NaCl} to the solution of \ce{CaCl_2} enhances the hedgroup 
%order parameter response compared to the solution with only \ce{CaCl_2}. \cite{altenbach84}
%\todo{Simulate this effect and discuss it further}

%\todo{The difference between DPPC and POPC -- simulate and compare with experiment. }

%\todo{It might be worth acknowledging each experimental finding in \cite{Altenbach84} and observations in \cite{Javanainen2017-ChemComm}.}

@@

\newpage
\newpage
\appendix

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

Zwitterionic phosphocholine (PC) lipid bilayers are commonly used model systems for
cellular membranes. Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in PC lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Interpretation most of these experiments suggests that Na$^+$ ions exhibit
negligible binding in PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules. Atomistic resolution
molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. This study presented 
a direct comparison between ion binding affinity between simulations and
experiments by using NMR data for lipid headgroup order 
parameters and the electrometer concept~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
As a starting point we use the parameters from the Lipid14 model \cite{dickson14},
which gave the best cation binding behaviour in the previous study \cite{catte16}.
The developed ECC-lipid parameters reproduce the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly for reproducing the membrane binding affinities and induced
structural effects of Sodium and Calcium ions.

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of the lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased back
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in the Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules.
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models.}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stronger electrostatic interactions with them.
After reducing the $\sigma$ parameters by a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
back in agreement with the experimental value (see Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding between experiments and simulations was compared by 
using lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid,  $X^\pm$~\cite{seelig87}.
Change in the order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

The change of the head group order parameters is empirically quantified
as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because recent analysis concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. This is rationalized as a change
of lipid head group dipole tilt toward water phase with bound positive
charge and {\it vice versa} with negative charge \cite{seelig87}.

@@

The measured order parameter change depends on the response of 
the headgroup on the bound charge, i.e. $m_i$ in Eq.~\ref{OPchangeEQ},
and the ion binding affinity. Thus, the former property has to be
well quantified before using the electrometer concept to
analyze binding affinities. This is done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response also in simulations,
we use experimental data for dihexadecyldimethylammonium (2C$_{16}$$^+$N2C$_1$Br$^-$)
in a POPC bilayer \cite{scherer89}. Dihexadecyldimethylammonium
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems \cite{scherer87}.

\subsection{Salt concentrations and binding affinity}
Experimental studies providing the lipid head group order
parameters report salt concentrations in two different ways.
For DPPC, salt concentrations before solvating
the lipids are used~\cite{akutsu81}. Alternatively, for POPC,
the salt concentration of supernatant after solvation is
measured using atomic absorption~\cite{altenbach84}.
\todo{JOE: I'm afraid that the above description might sound to the reader
  like for some reasons one method is used with DPPC, but a different one with POPC.
  I think simply writing one way is\dots another way is \dots (or similarly) will be less confusing.}
In this work, we will use the latter definition. For this we
will calculate the salt concentration in the aqueous bulk region,
\textit{i.e.}, the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the concentration was calculated as in the DPPC experiments.
Despite of the measurable differences between these two concentrations for CaCl$_2$ systems,
the qualitative conclusions are affected neither in this
nor in our previous work~\cite{catte16}.

To quantify the ion binding affinity to a membrane, we
calculated the relative surface excess of ions with respect to water $\Gamma_i^w$  \cite{chattorajBOOK}.
This quantity can be used to compare binding affinity between different models,
because it does not depend on the definition of the interfacial region
or the position of the Gibbs dividing plane between two bulk regions.
\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
Therefore, we consider the hydrophobic interior of a membrane having zero concentrations of ions and water
as a bulk region, which is separated by a monolayer interface from 
a another region with bulk concentration of ions in water.
Since the boundary of the latter region is set to the edge of simulation box,
all molecules in simulation are within the defined interfacial region. 
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

Lipid bilayer structure without ions was validated against NMR and 
\mbox{x-ray} scattering experiments by calculating order parameters
for C-H bonds and form factors from our simulations. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules by using Eq. \ref{OP}. Form factors were calculated 
from equation
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length in the bulk region,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning until the water bulk region, $D$.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
The simulated systems consisted of POPC bilayer in pure water
or in varying salt concentrations.
In particular, the periodic orthorhombic simulation box contained 
128 POPC molecules and approximately 50~water molecules per each lipid.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
presented in Supporting Information (SI). 
We used the ECC-ions model for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium \\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH2{)}_{15}-CH_3)(-[:30]{(}CH2{)}_{15}-CH_3)
} , 
\vspace{0.5cm} \\
which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
%Chloride counter ions were used
%in simulations, because parameters for bromide, which was used in the experiment \cite{scherer89},
%were not available in the standard Gromacs files for Amber force field.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.

@@

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
The dominant contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Such a finding is based on the calculated
number of contacts per lipid between \ce{Ca^{2+}} and POPC oxygen atoms, 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
in which a contact between the two groups is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ (robustness of this definition discussed in reference~\citenum{javanainen17}). 
the average number of contacts between \ce{ca^{2+}} and any oxygen atom in popc is 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.47 $,    % = 59.8/128 
whereas if only phosphate oxygen atoms are considered 
the average number of contacts decreases only by a tiny amount to 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.45 $,    % = 57.2/128 . 
and if only carbony oxygen atoms are considered,
this value is merely
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.02 $.    % = (59.8-57.2)/128 . 
We corroborate this finding with the probability isodensity contours in Fig.~\ref{fig:volmaps}, 
which are easier to interpret.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From a recent theoretical work with long simulations 
(Fig. S17 in \citenum{javanainen17} -- C36 with MK Ca2+, any other model, either different cation or Slipids/C36, just such calcium)
this time can be roughly estimated -- get the simulation from \url{https://doi.org/10.5281/zenodo.259376} and analyze it -- 
in the order of $1$--$10\,\mu\mathrm{s}$~\cite{javanainen17}. 
This is in contrast to our model, ECC-lipids, 
which estimates the residence time of calcium bound to a POPC membrane to be 
smaller than $53\,\mathrm{ns}$ for 90\% of the time a cation is in a membrane-bound state. 
i.e. at least two orders of magnitude lower than previous estimates.
In addition, the longest observed residence time in our simulation is $120\,\mathrm{ns}$, 
which is well below the length of our simulations.
This is unfortunately not the case of many theoretical works,
where the residence time cannot be estimated,
or the simulation has not even decorrelated from it's 
initial condition with a population of bound/unbound cations \cite{catte16}, 
because it is limited by the simulation length, 
even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
Increasing time scales to the order of $\mu\mathrm{s}$ 
doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used. 
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent. 
A histogram of residence times from our simulations is in Fig.~S\ref{fig:hist_residence_times} in SI. 
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
Add data for Na+. }

@@

\newpage
\appendix
